Reasonably Programmable Literal Notation

### 2018-06-18 icfp18/syntax-icfp18.tex
TR number and removed todo about github URL

\paragraph{Problem} The problem that specifically motivates this paper is that these existing systems  make it difficult or impossible to reason abstractly about such fundamental issues as types and variable binding when presented with a program using user-defined literal forms. Instead, programmers and editor services can only reason transparently, i.e. by inspecting the underlying expansion or the implementation details of the collection of extensions responsible for producing the expansion.% In some existing systems, even determining which syntax extension is responsible for a literal form is difficult.
%, i.e. when developing large programs ``consisting of many small programs o(modules), possibly written by different people''
\begin{figure}[t!]
\vspace{4px}
\begin{subfigure}[t]{0.46\textwidth}
\fcolorbox{gray!20}{gray!20}{\makebox[0.97\textwidth][l]{\textsc{existing grammar extension systems}}}
\begin{lstlisting}[xleftmargin=-2pt, morekeywords={EXTEND}]
  EXTEND /* loaded by, e.g., camlp4 */
    expr:
      |  "`(" q = kquery ")`" -> q
    kquery: 
      | /* ...K query grammar... */
  /* ...more extensions defined... */
  let x = compute_x();
  let y = `((!R)@&{&/x!/:2_!x}'!R)`;
\end{lstlisting}
\vspace{-5px}
\caption{It is difficult to reason abstractly given program text that uses a variety of grammar extensions (see the six reasoning criteria in the text).}
\label{fig:K-dialect}
\end{subfigure}
\hfill
\begin{subfigure}[t]{0.51\textwidth}
\fcolorbox{gray!20}{gray!20}{\makebox[\textwidth][l]{\textsc{\textbf{this paper}: typed literal macros}}}
\begin{lstlisting}[xleftmargin=-2pt]
  notation $kq at KQuery.t {
    lexer  KQueryLexer
    parser KQueryParser.start
    in package kquery_parser;
    dependencies = {module KQuery = KQuery}
  }; /* ...more notations defined... */
  let x = compute_x();
  let y = $kq `(SURL(!R)@&{&/EURLxSURL!/:2_!EURLxSURL}'!REURL)`;
\end{lstlisting}
\vspace{-5px}
\caption{TLMs make examples like these more reasonable by leaving the base grammar fixed and strictly enforcing a simple type, binding and segmentation discipline.}
\label{fig:K-tsm-example}
\end{subfigure}
\vspace{2px}
\caption{Two of the possible ways to introduce literal notation for encodings of K queries}
\vspace{-4px}
\end{figure}

@@

\paragraph{Paper Outline} 
% \begin{itemize}
% \item 
Sec.~\ref{sec:setlms} details expression TLMs in Reason with several more case studies of varying detail, notably including TLMs for regular expressions, HTML literals, chemical structure literals and quasiquotation for Reason language terms, which can be used for implementing other TLMs. It also describes experimental integration with MetaOCaml, which can help providers reason about type correctness.
% \item 
Sec.~\ref{sec:sptsms} then briefly introduces {pattern TLMs} and describes the special reasoning conditions in pattern position. %The hygiene condition for pattern TLMs is interesting.
% \item 
Sec.~\ref{sec:ptsms} introduces the more general {parametric TLMs}, which allow us to define literal notation at a type- or module-parameterized family of types.
 % This section also reveals that the \li{dependencies} clause can be understood as parameterization followed immediately by partial parameter application, or, alternatively, that parameterization can be understood in terms of the \li{dependencies} clause together with ML's module functions (functors). 
 Having introduced the basic machinery by example, we proceed in 
%Sec.~\ref{sec:more-examples} with examples that further demonstrate the expressive power of this approach and various parser implementation strategies. Notably, this section gives our take on the example from  Fig.~\ref{fig:urweb} of Ur/Web-style HTML literals as well as a TLM that implements quasiquotation for Reason language terms, which is useful for implementing other TLMs. 
Sec.~\ref{sec:setlms-formally} to describe a type-theoretic account of simple expression and pattern TLMs and formally establish the reasoning principles implied above in their essential form. The full technical details and proofs are in the accompanying technical report \cite{icfp18tr}.\todo{fix .bib when we have the TR number}{} 
% Sec.~\ref{sec:ptlms-formally} adds type functions and an ML-style module system to this calculus, and gives a more general variant of the reasoning principles theorem. 
Sec.~\ref{sec:implementation} provides a brief overview of how we are implementing TLMs for Reason without modifying OCaml's type system. % We also give a brief overview of optional integration with MetaOCaml, which can be used to guarantee that all expansion generated by a TLM are type correct. 
This implementation, called \li{Relit}, and additional implementation details, documentation and examples are available from the project page\todo{get a github.io url?}{}:

>>>>

\paragraph{Problem} The problem that specifically motivates this paper is that these existing systems  make it difficult or impossible to reason abstractly about such fundamental issues as types and variable binding when presented with a program using user-defined literal forms. Instead, programmers and editor services can only reason transparently, i.e. by inspecting the underlying expansion or the implementation details of the collection of extensions responsible for producing the expansion.% In some existing systems, even determining which syntax extension is responsible for a literal form is difficult.
%, i.e. when developing large programs ``consisting of many small programs o(modules), possibly written by different people''
\begin{figure}[t!]
\vspace{4px}
\begin{subfigure}[t]{0.46\textwidth}
\fcolorbox{gray!20}{gray!20}{\makebox[0.97\textwidth][l]{\textsc{existing grammar extension systems}}}
\begin{lstlisting}[xleftmargin=-2pt, morekeywords={EXTEND}]
  EXTEND /* loaded by, e.g., camlp4 */
    expr:
      |  "`(" q = kquery ")`" -> q
    kquery: 
      | /* ...K query grammar... */
  /* ...more extensions defined... */
  let x = compute_x();
  let y = `((!R)@&{&/x!/:2_!x}'!R)`;
\end{lstlisting}
\vspace{-5px}
\caption{It is difficult to reason abstractly given program text that uses a variety of grammar extensions (see the six reasoning criteria in the paper text).}
\label{fig:K-dialect}
\end{subfigure}
\hfill
\begin{subfigure}[t]{0.51\textwidth}
\fcolorbox{gray!20}{gray!20}{\makebox[\textwidth][l]{\textsc{\textbf{this paper}: typed literal macros}}}
\begin{lstlisting}[xleftmargin=-2pt]
  notation $kq at KQuery.t {
    lexer  KQueryLexer
    parser KQueryParser.start
    in package kquery_parser;
    dependencies = {module KQuery = KQuery}
  }; /* ...more notations defined... */
  let x = compute_x();
  let y = $kq `(SURL(!R)@&{&/EURLxSURL!/:2_!EURLxSURL}'!REURL)`;
\end{lstlisting}
\vspace{-5px}
\caption{TLMs make examples like these more reasonable by leaving the base grammar fixed and strictly enforcing a simple type, binding and segmentation discipline.}
\label{fig:K-tsm-example}
\end{subfigure}
\vspace{2px}
\caption{Two of the possible ways to introduce literal notation for encodings of K queries}
\vspace{-4px}
\end{figure}

@@

\paragraph{Paper Outline} 
% \begin{itemize}
% \item 
Sec.~\ref{sec:setlms} details expression TLMs in Reason with several more case studies of varying detail, notably including TLMs for regular expressions, HTML literals, chemical structure literals and quasiquotation for Reason language terms, which can be used for implementing other TLMs. It also describes experimental integration with MetaOCaml, which can help providers reason about type correctness.
% \item 
Sec.~\ref{sec:sptsms} then briefly introduces {pattern TLMs} and describes the special reasoning conditions in pattern position. %The hygiene condition for pattern TLMs is interesting.
% \item 
Sec.~\ref{sec:ptsms} introduces the more general {parametric TLMs}, which allow us to define literal notation at a type- or module-parameterized family of types.
 % This section also reveals that the \li{dependencies} clause can be understood as parameterization followed immediately by partial parameter application, or, alternatively, that parameterization can be understood in terms of the \li{dependencies} clause together with ML's module functions (functors). 
 Having introduced the basic machinery by example, we proceed in 
%Sec.~\ref{sec:more-examples} with examples that further demonstrate the expressive power of this approach and various parser implementation strategies. Notably, this section gives our take on the example from  Fig.~\ref{fig:urweb} of Ur/Web-style HTML literals as well as a TLM that implements quasiquotation for Reason language terms, which is useful for implementing other TLMs. 
Sec.~\ref{sec:setlms-formally} to describe a type-theoretic account of simple expression and pattern TLMs and formally establish the reasoning principles implied above in their essential form. The full technical details and proofs are in the accompanying technical report \cite{icfp18tr}.
% Sec.~\ref{sec:ptlms-formally} adds type functions and an ML-style module system to this calculus, and gives a more general variant of the reasoning principles theorem. 
Sec.~\ref{sec:implementation} provides a brief overview of how we are implementing TLMs for Reason without modifying OCaml's type system. % We also give a brief overview of optional integration with MetaOCaml, which can be used to guarantee that all expansion generated by a TLM are type correct. 
This implementation, called \li{Relit}, and additional implementation details, documentation and examples are available from the project page:
